Arrows for Parallel Computations

### 2018-01-05 paper/content/arrows.tex
wip
own remarks until page 17.

\subsection{Arrows}
\label{sec:arrows}
Arrows were introduced by \citet{HughesArrows} as a general interface for computation and a less restrictive generalisation of Monads. \citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator |(>>=) :: m a -> (a -> m b) -> m b| (with |m| being a Monad). In the example a parser of the type |Parser s a| with static meta information |s| and result |a| is shown to not be able to use the static information |s| without applying the monadic function |a -> m b|. With Arrows this is possible.

@@

In this paper we will show that we can express parallel computations with Arrows and do not require Monads to do so. We also do not restrict the compatible Arrows to ones which have |ArrowApply| instances but instead only require instances for |ArrowChoice| and |ArrowLoop|. Because of this, we have a truly more general interface as compared to a monadic one.

Also note that, while we could have based our DSL on Profunctors as well, we chose Arrows. for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability. However, they are a promising candidate for future improvements of our DSL. Note that some Profunctors, especially ones supporting a composition operation among other things, can already be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.

>>>>

\subsection{Arrows}
\label{sec:arrows}
Arrows were introduced by \citet{HughesArrows} as a general interface for computation and a less restrictive generalisation of Monads. \citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator |(>>=) :: m a -> (a -> m b) -> m b| (with |m| being a Monad)\footnote{In the example a parser of the type |Parser s a| with static meta information |s| and result |a| is shown to not be able to use the static information |s| without applying the monadic function |a -> m b|. With Arrows this is possible.}.

@@

In this paper we will show that parallel computations can be expressed with this more general interface of Arrows without requiring Monads. We also do not restrict the compatible Arrows to ones which have |ArrowApply| instances but instead only require instances for |ArrowChoice| (for if-then-else constructs) and |ArrowLoop| (for looping). Because of this, we have a truly more general interface as compared to a monadic one.

Also note that, while we could have based our DSL on Profunctors as well, we chose Arrows \mbcomment{cite missing?}. for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability. However, they are a promising candidate for future improvements of our DSL. Note that some Profunctors, especially ones supporting a composition operation among other things, can already be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.

### 2018-01-05 paper/content/basicSkeletons.tex
own remarks until page 17.

\subsection{Extending the Interface}
\label{sec:extending-interface}
With the |ArrowParallel| type class in place and implemented, we can now implement some further basic parallel interface functions. These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.

\subsubsection{Lazy |parEvalN|}
\begin{figure}[tb]
	\includegraphics[scale=0.7]{images/parEvalNLazy}
	\caption{Schematic depiction of |parEvalNLazy|.}
	\label{fig:parEvalNLazyImg}
\end{figure}
The function |parEvalN| is 100\% strict, which means that it fully evaluates all passed arrows. Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g. |map (arr . (+)) [1..]| or just because we need the arrows evaluated in chunks. |parEvalNLazy| (Figs.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this. It works by first chunking the input from |[a]| to |[[a]]| with the given |ChunkSize| in |arr (chunksOf chunkSize)|. These chunks are then fed into a list |[arr [a] [b]]| of parallel arrows created by feeding chunks of the passed |ChunkSize| into the regular parEvalN by using |listApp|. The resulting |[[b]]| is lastly converted into |[b]| with |arr concat|.
\begin{figure}[t]
\begin{code}
parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
parEvalNLazy conf chunkSize fs =
	arr (chunksOf chunkSize) >>>
    evalN fchunks >>>
    arr concat
    where
      fchunks = map (parEvalN conf) (chunksOf chunkSize fs)
\end{code} %$ %% formatting
\caption{Definition of |parEvalNLazy|.}
\label{fig:parEvalNLazy}
\end{figure}

\subsubsection{Heterogeneous tasks}
\begin{figure}[tb]
	\includegraphics[scale=0.7]{images/parEval2}
	\caption{Schematic depiction of |parEval2|.}
	\label{fig:parEval2Img}
\end{figure}
We have only talked about the parallelisation of arrows of the same type until now. But sometimes we want to parallelize heterogeneous types as well. We can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| type class. The idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into |parEvalN|.

>>>>

\subsection{Extending the Interface}
\label{sec:extending-interface}
With the |ArrowParallel| type class in place and implemented, we can now define some further basic parallel interface functions. These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.

\subsubsection{Lazy |parEvalN|}
\begin{figure}[tb]
	\includegraphics[scale=0.7]{images/parEvalNLazy}
	\caption{Schematic depiction of |parEvalNLazy|.}
	\label{fig:parEvalNLazyImg}
\end{figure}
|parEvalN| fully traverses the list of passed Arrows as well as their inputs. Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g. |map (arr . (+)) [1..]| or just because we need the arrows evaluated in chunks. |parEvalNLazy| (Figs.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this. It works by first chunking the input from |[a]| to |[[a]]| with the given |chunkSize| in |arr (chunksOf chunkSize)|. These chunks are then fed into a list |[arr [a] [b]]| of chunk-wise parallel Arrows with the help of our lazy and sequential |evalN|. The resulting |[[b]]| is lastly converted into |[b]| with |arr concat|.
\begin{figure}[t]
\begin{code}
parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
parEvalNLazy conf chunkSize fs =
	arr (chunksOf chunkSize) >>>
    evalN fchunks >>>
    arr concat
    where
      fchunks = map (parEvalN conf) (chunksOf chunkSize fs)
\end{code} %$ %% formatting
\caption{Definition of |parEvalNLazy|.}
\label{fig:parEvalNLazy}
\end{figure}

\subsubsection{Heterogeneous tasks}
\begin{figure}[tb]
	\includegraphics[scale=0.7]{images/parEval2}
	\caption{Schematic depiction of |parEval2|.}
	\label{fig:parEval2Img}
\end{figure}
We have only talked about the parallelisation of arrows of the same set of input and output types until now. But sometimes we want to parallelize heterogeneous types as well. We can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| type class. Here, the general idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into |parEvalN|.

### 2018-01-05 paper/content/futures.tex
own remarks until page 17.

While the above example could be rewritten into only one |parEvalN| call by directly wiring the arrows together before spawning, it illustrates an important problem. When using a |ArrowParallel| backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}. This can become a serious bottleneck
for larger amount of data and number of processes \citep[as e.g.][showcases]{Berthold2009-fft}.
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
	\caption[without Futures]{Communication between 4 Eden processes without Futures. All communication goes through the master node. Each bar represents one process. Black lines between processes represent communication. Colours: blue $\hat{=}$ idle, green $\hat{=}$ running, red  $\hat{=}$ blocked, yellow $\hat{=}$ suspended.}
	\label{fig:withoutFutures}
% \olcomment{more practical and heavy-weight example! fft (I have the code)?}
% \mbcomment{Depends... Are the communications easy to read in such an example?}
% \mbcomment{Keep the description for the different colours, or link to
%   the EdenTV description in \ref{sec:edentv}}
% \olcomment{ok as is}
% \olcomment{use the fft example (when it works)?}
\end{figure}

@@

In our communication example we can use this |Future| concept for direct communication between nodes as shown in Fig.~\ref{fig:someCombinatorParallel}. Note that with our PArrows DSL we can also define default instances |Future fut a ()| for each backend similar to how |ArrowParallel arr a b ()| was defined in Section~\ref{sec:parallel-arrows}. Details can be found in Section~\ref{app:omitted}.
\begin{figure}[tbh]
\begin{code}
someCombinator :: (ArrowChoice arr, NFData b, NFData c,
	ArrowParallel arr a (fut b) (), 
	ArrowParallel arr (fut b) c (),
	Future fut b ()) =>
	[arr a b] -> [arr b c] -> arr [a] [c]
someCombinator fs1 fs2 =
	parEvalN () (map (>>> put ()) fs1) >>>
	rightRotate >>>
	parEvalN () (map (get () >>>) fs2)
\end{code}
\caption{The mock-up combinator in parallel.}
\label{fig:someCombinatorParallel}
\end{figure}
In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed -- similar to what is shown in the trace visualisation in Fig.~\ref{fig:withFutures}. One especially elegant aspect of this solution is that we can specify the type of |Future| to be used per backend with full interoperability between code using different backends, without even requiring to know about the actual type used for communication. We only specify that there has to be a compatible Future and do not care about any specifics. \olcomment{Fig. is not really clear. Do Figs with a lot of load? --- fft?}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withFutures}
	\caption[with Futures]{Communication between 4 Eden processes with Futures. Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (black lines between the bars, one example message is marked in the Figure) instead of always going through the master node (bottom bar).}
	\label{fig:withFutures}
\end{figure}

>>>>

While the above example could be rewritten into only one |parEvalN| call by directly wiring the arrows together before spawning, it illustrates an important problem. When using a |ArrowParallel| backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}. This can become a serious bottleneck
for a larger amount of data and number of processes \citep[as e.g.][showcases]{Berthold2009-fft}.
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
	\caption[without Futures]{Communication between 4 Eden processes without Futures. All communication goes through the master node. Each bar represents one process. Black lines between processes represent communication. Colours: blue $\hat{=}$ idle, green $\hat{=}$ running, red  $\hat{=}$ blocked, yellow $\hat{=}$ suspended.}
	\label{fig:withoutFutures}
% \olcomment{more practical and heavy-weight example! fft (I have the code)?}
% \mbcomment{Depends... Are the communications easy to read in such an example?}
% \mbcomment{Keep the description for the different colours, or link to
%   the EdenTV description in \ref{sec:edentv}}
% \olcomment{ok as is}
% \olcomment{use the fft example (when it works)?}
\end{figure}

@@

In our communication example we can use this |Future| concept for direct communication between nodes as shown in Fig.~\ref{fig:someCombinatorParallel}.
\begin{figure}[tbh]
\begin{code}
someCombinator :: (ArrowChoice arr, NFData b, NFData c,
	ArrowParallel arr a (fut b) (), 
	ArrowParallel arr (fut b) c (),
	Future fut b ()) =>
	[arr a b] -> [arr b c] -> arr [a] [c]
someCombinator fs1 fs2 =
	parEvalN () (map (>>> put ()) fs1) >>>
	rightRotate >>>
	parEvalN () (map (get () >>>) fs2)
\end{code}
\caption{The mock-up combinator in parallel.}
\label{fig:someCombinatorParallel}
\end{figure}
In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed -- similar to what is shown in the trace visualisation in Fig.~\ref{fig:withFutures}. One especially elegant aspect of the definition in Fig.~\ref{fig:future} is that we can specify the type of |Future| to be used per backend with full interoperability between code using different backends, without even requiring to know about the actual type used for communication. We only specify that there has to be a compatible Future and do not care about any specifics as can be seen in Fig.~\ref{fig:someCombinatorParallel}. Note that with our PArrows DSL we can also define default instances |Future fut a ()| for each backend similar to how |ArrowParallel arr a b ()| was defined in Section~\ref{sec:parallel-arrows}. Details can be found in Section~\ref{app:omitted}. \olcomment{Fig. is not really clear. Do Figs with a lot of load? --- fft?}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withFutures}
	\caption[with Futures]{Communication between 4 Eden processes with Futures. Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (black lines between the bars, one example message is marked in the Figure) instead of always going through the master node (bottom bar).}
	\label{fig:withFutures}
\end{figure}

### 2018-01-05 paper/content/mapSkeletons.tex
own remarks until page 17.

\section{Skeletons}
\label{sec:skeletons}
Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations. While there are many possible skeletons to implement, we regard in detail here only some |map|-based and topological skeletons to demonstrate the power of PArrows.
%%% \FloatBarrier
\subsection{|map|-based Skeletons}
\label{sec:map-skeletons}
We start with |map|-based skeletons. The essential differences between the skeletons presented here are in terms of order of evaluation and work distribution but still provide the same output of a standard |map|.

\paragraph{Parallel |map| and laziness.}
The |parMap| skeleton (Figs.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs. We can implement it with |ArrowParallel| by repeating an arrow |arr a b| and then passing it into |parEvalN| to obtain an arrow |arr [a] [b]|.
Just like |parEvalN|, |parMap| is 100\% strict.
Because of this, it has the same restrictions as |parEvalN| as compared to |parEvalNLazy|. So it makes sense to also have a |parMapStream| (Figs.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like |parMap|, but uses |parEvalNLazy| instead of |parEvalN|. The code of these two skeletons is quite straightforward, we show it in Appendix.

@@

\paragraph{Statically load-balancing parallel |map|.}
Our |parMap| spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we gave in this paper). This can be quite wasteful and a statically load-balancing |farm| (Figs.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers seems useful.
The definitions of the helper functions |unshuffle|, |takeEach|, |shuffle| (Fig.~\ref{fig:edenshuffleetc}) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.

%\paragraph{Lazy statically load-balancing parallel map}
Since a |farm|  is basically just |parMap| with a different work distribution, it is, again, 100\% strict. So we can define |farmChunk| (Figs.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses |parEvalNLazy| instead of |parEvalN|. It is basically the same definition as for |farm|, with |parEvalN| replaced with |parEvalNLazy|.

>>>>

\section{Skeletons}
\label{sec:skeletons}
Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations. While there are many possible skeletons to implement, here we only regard in detail some |map|-based and topological skeletons to demonstrate the power of PArrows.
%%% \FloatBarrier
\subsection{|map|-based Skeletons}
\label{sec:map-skeletons}
We start with |map|-based skeletons. The essential differences between the skeletons presented here are in terms of order of evaluation and work distribution but still provide the same output of a standard |map|.

\paragraph{Parallel |map| and laziness.}
The |parMap| skeleton (Figs.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs. We can implement it with |ArrowParallel| by repeating an arrow |arr a b| and then passing it into |parEvalN| to obtain an arrow |arr [a] [b]|.
Just like |parEvalN|, |parMap| traverses all input Arrows as well as the inputs.
Because of this, it has the same restrictions as |parEvalN| as compared to |parEvalNLazy|. So it makes sense to also have a |parMapStream| (Figs.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like |parMap|, but uses |parEvalNLazy| instead of |parEvalN|. The code of these two skeletons is quite straightforward, we show it in Appendix \ref{app:omitted} in Figs.\ref{fig:parMap} and \ref{fig:parMapStream}.

@@

\paragraph{Statically load-balancing parallel |map|.}
Our |parMap| spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we presented in this paper). This can be quite wasteful and a statically load-balancing |farm| (Figs.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers seems useful.
The definitions of the helper functions |unshuffle|, |takeEach|, |shuffle| (Fig.~\ref{fig:edenshuffleetc}) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.

%\paragraph{Lazy statically load-balancing parallel map}
Since a |farm|  is basically just |parMap| with a different work distribution, it has the same restrictions as |parEvalN| and |parMap|. We can, however, define |farmChunk| (Figs.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses |parEvalNLazy| instead of |parEvalN|. It is basically the same definition as for |farm|, but with |parEvalNLazy| instead of |parEvalN|.

### 2018-01-05 paper/content/motivation.tex
own remarks until page 17.

Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms including the expression of parallelism. Haskell, which we focus on in this paper, has  several mature implementations. We regard here in-depth
Glasgow parallel Haskell or short GpH (its Multicore SMP implementation, in particular), the
|Par| Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a monad, even if
only for the internal representation. Some introduce additional
language constructs. Section \ref{sec:parallelHaskells} gives a short overview over these languages.

@@

We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner -- we tame the zoo of parallel Haskells. We provide a
common, very low-penalty programming interface that allows to switch
the parallel backends at will. The induced penalty was in the single-digit percent range, with means over the varying cores configuration typically under 1\% overhead in our measurements (Section~\ref{sec:benchmarks}). Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.

>>>>

Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms. Haskell, which we focus on in this paper, has  several mature implementations. We regard here in-depth
Glasgow parallel Haskell or short GpH (its Multicore SMP implementation, in particular), the
|Par| Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a monad, even if
only for the internal representation. Some introduce additional
language constructs. Section \ref{sec:parallelHaskells} gives a short overview over these languages.

@@

We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner -- we tame the zoo of parallel Haskells. We provide a
common, very low-penalty programming interface that allows to switch
the parallel backends at will. The induced penalty was in the single-digit percent range, with means over the varying cores configuration typically under 1\% overhead in our measurements (Section~\ref{sec:benchmarks}). Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.

### 2018-01-05 paper/content/parallelHaskells.tex
own remarks until page 17.

In the above definition of |parEvalN| we just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|. % $
We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator. We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evaluates to normal form.
%
Fig.~\ref{fig:parEvalNMulticoreImg} shows a visual representation of this code.

@@

\subsubsection{Eden}
Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with a MPI and a PVM backends.\footnote{The projects homepage can be found at \url{http://www.mathematik.uni-marburg.de/~eden/}. The Hackage page is at \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}.} It is targeted towards clusters, but also functions well in a shared-memory setting with a further simple backend. However, in contrast to many other parallel Haskells, in Eden each process has its own heap. This seems to be a waste of memory, but with distributed programming paradigm and individual GC per process, Eden yields good performance results on multicores, as well \cite{arcs-dc,aswad2009low}.

>>>>

In the above definition of |parEvalN| we just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|. % $
We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator. We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evaluates to normal form. Other strategies like e.g. evaluation to weak head normal form are available as well. It also allows for custom |Strategy| implementations to be used.
%
Fig.~\ref{fig:parEvalNMulticoreImg} shows a visual representation of this code.

@@

\subsubsection{Eden}
Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with MPI and PVM as distributed backends.\footnote{The projects homepage can be found at \url{http://www.mathematik.uni-marburg.de/~eden/}. The Hackage page is at \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}.} It is targeted towards clusters, but also functions well in a shared-memory setting with a further simple backend. However, in contrast to many other parallel Haskells, in Eden each process has its own heap. This seems to be a waste of memory, but with distributed programming paradigm and individual GC per process, Eden yields good performance results on multicores, as well \cite{arcs-dc,aswad2009low}.
